Linear regression for approximating real and synthetic data

### 2022-10-10 doc/paper.tex
Update paper

%%% use twocolumn and 10pt options with the asme2ej format
\documentclass[twocolumn,10pt,cleanfoot]{asme2ej}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{\it We use Ordinary Least Squares, Ridge and Lasso regression on a synthetic dataset generated by the Franke Function, and on a real dataset of elevation data from a terrain in Norway. We explore different regularization values and polynomial degrees and find that that all methods perform similarly on both sets of data. We find that linear regression performs well on the synthetic data, but falls short when approximating elevation data with large variances in height.
}
\end{abstract}

@@

Linear regression is a simple method in machine learning, used to create a fit of a dataset of which it can produce continuous predictions. This is achieved by assuming that we can model the output values as a linear combination of continuous functions of the input values. For example, we can assume that the observed values in our dataset can be approximated by different polynomial degrees of the input values and find the linear coefficients $\beta$ which minimize the difference between our observed and predicted values. This general idea that we can create a model which approximates our observed data with a minimal difference between observed and predicted values is a problem at the very heart of machine learning.

However, in order to create an optimal model, we require optimal parameters which cannot be known a priori. Thus, we require extensive testing and tuning of parameters to inform our selection. This presents a major obstacle, as we must for example find the optimal polynomial degree $N$ for which our model best approximates the observed values at the same time as we tune the hyperparameter $\lambda$ responsible for an adequate regularisation. We must find these parameters which lie at an intersection, where all parameters yield the minimal error in unity.

Furthermore, we will compare different models of linear regression, namely Ordinary Least Squares (OLS), Ridge and Lasso regression, and employ resampling methods, namely bootstrap and cross-validation to ensure that we draw conclusive results as these help us reduce the randomness in our parameters based off how we split our data. <source?>

@@

This sections covers the theoretic background which underlies the methods used in this paper.

@@

The correctness of our linear regression can be confirmed by running our debug function (the so called SkrankeFunction)\footnote{this can be done by running \texttt{\$ python3 task\_b.py --debug -n 2}}, which is a second degree polynomial where $b_0 = 0, \; b_1 = 1, \; b_2 = 2$ and so on.

@@

Using Ridge regression, we see from Fig.~\ref{frankiedifflambdas} that larger and larger values of $\lambda$ efficiently prevents overfitting and reduces the variance between predictions. Looking at the beta progression with for the same values of $\lambda$ (Fig.~\ref{frankeridgebetas}), we see that the regularization parameter is successfully shrinking the beta values. Using Scikit's gridsearch, we find the optimal $\lambda$ to be $10^{-8}$ and the optimal polynomial degree to be 7, with an average MSE of $0.0037999$, which is very slightly lower than OLS for the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$. We see from this that Ridge regression has not really had much of an effect in decreasing the MSE. We must now remember what Ridge aims to do. With the regularization parameter, we can prevent overfitting if our model is too complex for our data. But we should recall that we decide the complexity of our model, by deciding which polynomial degree we use when creating the design matrix. In fact, by choosing polynomial degree 7 for our OLS method to get the best MSE for noisy data, we are in fact doing a primitive form of best subset selection, as described in Hastie et Al.\cite[57]{hastie}. Almost by definition, the polynomial degree with the best MSE is not overfitted, because if it was, it would not be the model with the lowest MSE among all polynomial degrees. Therefore it is not so unexpected that Ridge regression has a relatively low impact on the MSE. We could expect that Ridge would allow us to reach higher polynomial degrees without overfitting, by shrinking features of the higher polynomial degrees that are not supported but allowing us to keep other high degree terms (for example, allowing us to use $x^5y^4$, but discard $x^9$, if only $x^5y^4$ was supported by the data). But we can see from the Table. ~\ref{ols_vs_ridge_table} that we are not able to push into higher degrees. A possible explanation for this is that every feature is useful, due to the polynomial expansion of the Franke function.

@@

\subsection{Elevation data}

@@

Here we find some limitations of linear regression. We see from Fig. NUMBER that the we are able to fit the terrain up to a certain point, gaining an MSE of NUMBER with OLS with polynomial degree DEGREE. However, we see that beyond this, we have little to gain. This can be seen when we increase the polynomial degree to very large amounts, without seeing any decrease in MSE (Fig. NUMBER). Looking at the beta progression for selected betas (Fig. NUMBER), we see that the parameters stabilize, which implies that we are not overfitting, but simply reaching our best possible prediction for polynomial regression. The fact that we are not overfitting is further reinforced by looking at the bias-variance plot in Fig. NUMBER, which shows no increase in variance. It is reasonable to expect that our model will not overfit, due to the large increase in data we now use for the fit. Our synthetic data used $20 x 20 = 400$ datapoints, while we now have $320 x 640 = 204 800$ datapoints. This means that for a 50th order polynomial, which will have $1326$ features, our number of features are less than $1\%$ of our datapoints. If we decrease the number of datapoints used for training considerably, we can see overfitting start to occur (see Appendix 1). So we are not overfitting, but we are not decreasing our MSE either. This is likely because the elevation data has a jagged and complex shape, which cannot be accurately recreated, even with a polynomial of degree $50$. We should remember that the Franke function can be expanded into a series of polynomials, which is why we can expect to get a lower and lower MSE as the polynomial degree increases (so long as we have enough data and it is not noisy). The same is not true for our elevation data, and as such we have little to gain from increasing our model complexity after a certain point. Clearly, linear regression does not offer a very good tradeoff between computational cost and performance for this data.

@@

\begin{figure*}
% \begin{figure}[t]
\centerline{\includegraphics[width=3.25in]{figure/frankienoisecomparison.png}}
\caption{Test and train MSE for the Franke function with an increasing amount of noise}
\label{frankienoisecomparison}
\end{figure*}

>>>>

\documentclass[twocolumn,10pt,cleanfoot]{asme2ej}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Abstract}
We used linear regression to create a fit of a synthetic dataset generated by the Franke Function, and for real datasets of elevation data from terrain in Norway. More specifically, we used Ordinary Least Squares (OLS), Ridge and Lasso regression to explore how the different methods perform on different datasets. Comparing the Mean Squared Error (MSE) for the different methods, we found that Ridge and Lasso outperform OLS for small datasets and high polynomial degrees, but that OLS, given the optimal polynomial degree for the dataset, yields the lowest MSE. Overall we found that the different forms of linear regression fit the synthetic data well. However, when looking at the real datasets, we found that all forms of linear regression performed relatively poorly. We concluded that linear regression perhaps is not the best way of predicting terrain data with rapid and irregular changes in elevation.

@@

% One of the core challenges of machine learning is model selection, because we cannot make a priori assessments about how well the model will fit our dataset. The model is at the heart of machine learning, its like the CPU of the "machine" in machine learning. It really puts the "machine" in machine learning, so to speak. Learning is like the RAM. We are the keyboard and you are the screen.
Machine learning is well suited for finding complex relationships that are impossible to describe as a series of logical statements, or program by hand. Machine learning methods can look at millions of datapoints and parameters at once, and thus find links in the data which could never be found by manual inspection. But to do this, we require a model that is suited to our dataset. However, in order to create an optimal model, we require optimal parameters which cannot be known a priori. Thus, we require extensive testing and tuning of parameters to inform our selection. In this paper, we explore

% One of the core challenges of machine learning is model selection, because we cannot make a priori assessments about how well the model will fit our dataset. If our model is not complex enough for our problem, we are unlikely to reach a good prediction. Likewise, if our model is too complex, we run the risk of overfitting, where our model adapts to our training data perfectly and thus loses the ability to make predictions on unseen data. Given the enormous datasets that

% Linear regression is a simple method in machine learning, used to create a fit of a dataset of which it can produce continuous predictions. This is achieved by assuming that we can model the output values as a linear combination of continuous functions of the input values. For example, we can assume that the observed values in our dataset can be approximated by different polynomial degrees of the input values and find the linear coefficients $\beta$ which minimize the difference between our observed and predicted values. This general idea that we can create a model which approximates our observed data with a minimal difference between observed and predicted values is a problem at the very heart of machine learning.

% Furthermore, we will compare different models of linear regression, namely Ordinary Least Squares (OLS), Ridge and Lasso regression, and employ resampling methods, namely bootstrap and cross-validation to ensure that we draw conclusive results as these help us reduce the randomness in our parameters based off how we split our data. <source?>

@@

This sections covers the theoretic background which underlies the methods used in this paper. We use the following shorthands for our mathematical notation.

@@

The correctness of our linear regression can be confirmed by running our debug function (the so called SkrankeFunction)\footnote{this can be done by running \texttt{\$ python3 task\_b.py --debug -n 2 --noscale}}, which is a second degree polynomial where $b_0 = 0, \; b_1 = 1, \; b_2 = 2$ and so on. The SkrankeFunction (taken from \texttt{src/utils.py}) looks like this

@@

Using Ridge regression, we see from Fig.~\ref{frankiedifflambdas} that larger and larger values of $\lambda$ efficiently prevents overfitting and reduces the variance between predictions. Looking at the beta progression with for the same values of $\lambda$ (Fig.~\ref{frankeridgebetas}), we see that the regularization parameter is successfully shrinking the beta values. Using Scikit's gridsearch, we find the optimal $\lambda$ to be $10^{-8}$ and the optimal polynomial degree to be 7, with an average MSE of $0.0037999$, which is very slightly lower than OLS for the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$. We see from this that Ridge regression has not really had much of an effect in decreasing the MSE. We must now remember what Ridge aims to do. With the regularization parameter, we can prevent overfitting if our model is too complex for our data. Looking at Fig.~\ref{frankeridgenoisecomparison} in comparison to Fig.~\ref{frankienoisecomparison} (both in the appendix), we can see that Ridge really has achieved this, reducing overfitting as our data becomes more and more noisy. But we should recall that we decide the complexity of our model, by deciding which polynomial degree we use when creating the design matrix. In fact, by choosing polynomial degree 7 for our OLS method to get the best MSE for noisy data, we are in fact doing a primitive form of best subset selection, as described in Hastie et Al.\cite[57]{hastie}. Almost by definition, the polynomial degree with the best MSE is not overfitted, because if it was, it would not be the model with the lowest MSE among all polynomial degrees. Therefore it is not so unexpected that Ridge regression has a relatively low impact on the MSE. We could expect that Ridge would allow us to reach higher polynomial degrees without overfitting, by shrinking features of the higher polynomial degrees that are not supported but allowing us to keep other high degree terms (for example, allowing us to use $x^5y^4$, but discard $x^9$, if only $x^5y^4$ was supported by the data). But we can see from the Table. ~\ref{ols_vs_ridge_table} that we are not able to push into higher degrees. A possible explanation for this is that every feature is useful, due to the polynomial expansion of the Franke function.

@@

\subsection{Real elevation data}

@@

Here we find some limitations of linear regression. We see from Fig.~\ref{realmse} that the we are able to fit the terrain up to a certain point, gaining an MSE of NUMBER with OLS with polynomial degree DEGREE. However, we see that beyond this, we have little to gain. This can be seen when we increase the polynomial degree to very large amounts, without seeing any decrease in MSE (Fig. NUMBER). Looking at the beta progression for selected betas (Fig. NUMBER), we see that the parameters stabilize, which implies that we are not overfitting, but simply reaching our best possible prediction for polynomial regression. The fact that we are not overfitting is further reinforced by looking at the bias-variance plot in Fig. NUMBER, which shows no increase in variance. It is reasonable to expect that our model will not overfit, due to the large increase in data we now use for the fit. Our synthetic data used $20 x 20 = 400$ datapoints, while we now have $320 x 640 = 204 800$ datapoints. This means that for a 50th order polynomial, which will have $1326$ features, our number of features are less than $1\%$ of our datapoints. If we decrease the number of datapoints used for training considerably, we can see overfitting start to occur (see Appendix 1). So we are not overfitting, but we are not decreasing our MSE either. This is likely because the elevation data has a jagged and complex shape, which cannot be accurately recreated, even with a polynomial of degree $50$. We should remember that the Franke function can be expanded into a series of polynomials, which is why we can expect to get a lower and lower MSE as the polynomial degree increases (so long as we have enough data and it is not noisy). The same is not true for our elevation data, and as such we have little to gain from increasing our model complexity after a certain point. Clearly, linear regression does not offer a very good tradeoff between computational cost and performance for this data.

% \begin{figure} 
% \centerline{\includegraphics[width=3.25in]{figure/realmse.png}}
% \caption{Test and train MSE }
% \label{frankiedifflambdas}
% \end{figure}

@@

\begin{figure*}
% \begin{figure}[t]
\centerline{\includegraphics[width=3.25in]{figure/frankienoisecomparison.png}}
\caption{Test and train MSE for the OLS prediction of Franke function with an increasing amount of noise}
\label{frankienoisecomparison}
\end{figure*}

\begin{figure*}
% \begin{figure}[t]
\centerline{\includegraphics[width=3.25in]{figure/frankeridgenoisecomparison.png}}
\caption{Test and train MSE for the Ridge prediction with $\lambda = 10^{-8}$ of the Franke function with an increasing amount of noise}
\label{frankeridgenoisecomparison}
\end{figure*}
